Evolution Reinforces Cooperation with the Emergence of Self-Recognition Mechanisms: an empirical study of the Moran process for the iterated Prisoner's dilemma using reinforcement learning

### 2017-07-16 main.tex
Some additions and edits to the discussion and conclusion.

The strategies trained to maximize payoff in head-to-head matches are generally
cooperative and are effective invaders.
Combined with the fact that handshaking strategies are stronger resisters,
this suggests that while maximizing individual payoff can lead to the evolution
of cooperation, these strategies are not the most evolutionarily stable
in the long run. A strategy with a handshaking mechanism is still capable of
invading and is more resistant to subsequent invasions. Moreover, the
best resistor of the payoff maximally trained strategies (Evolved Looker Up
1\_1\_1),
which always defects if the opponent defects in the first round, is effectively
employing a one-shot handshake of C. These insights also suggest that a strategy
aware of the population distribution could choose to become a handshaker at
a critical threshold and use a different strategy for invasion when in the
minority. Information about the population distribution was not available
to our strategies.

We did not attempt other objective functions that may serve to select for both
invasion and resistance better than training at a starting population of
$(N/2, N/2)$. Nevertheless our results suggest that there is not much room for
improvement. Any handshake more sophisticated than always cooperate necessarily involves
a defection. (A handshake consisting of a long sequence of cooperations is
effectively a grudger.) For TF3 or EvolvedLookerUp1\_1\_1 to become better resistors
they need a longer or more strict handshake. But if this handshake involves
a defection then likely the invasion ability is diminished for $N > 2$: the top
invaders for larger $N$ are nice strategies that do not defect before their
opponents. This is because good invaders need to maximize match payoff to benefit
from fitness proportionate selection,
and so in the absence of a handshake mechanism, knowledge of the population
distribution, or some identifying label on the opponent,
a strategy must be generally cooperative. Aggressive strategies
are only effective invaders for the smallest $N$, dropping dramatically
as the population size increases.

We did, however, attempt to evolve CS using FSM and lookup table based players,
which resulted in some very similar strategies. In particular we evolved a
lookup strategy that had a handshake of DC and played TFT with other players
after a correct handshake while defecting otherwise, which is quite close in
function to CS (full grudging is not possible with a lookup table).

@@

A detailed empirical analysis of 164 strategies of the IPD within a pairwise
Moran process has been carried out. All \(\binom{164}{2}=13,366\) possible
ordered pairs of strategies have been placed in a Moran process with different
starting values allowing the each strategy to attempt to invade the other.
This is the largest such experiment carried out and has lead to many insights.

@@

Memory one strategies do not perform well in general in this study. There are
no memory one strategies in the top 5 performing strategies
for \(N>3\). This is due at least partly to their lack of ability to
recognise their opponents. More sophisticated strategies
prove to be high performers for invasion: these are infinite memory strategies
which have been trained using a number of reinforcement learning algorithms.
Interestingly they have been trained to perform well in tournaments and not
Moran processes which highlights the potential for improvement.

>>>>

The strategies trained to maximize payoff in head-to-head matches are generally
cooperative and are effective invaders.
Combined with the fact that handshaking strategies are stronger resisters,
this suggests that while maximizing individual payoff can lead to the evolution
of cooperation, these strategies are not the most evolutionarily stable
in the long run. A strategy with a handshaking mechanism is still capable of
invading and is more resistant to subsequent invasions. Moreover, the
best resistor of the payoff maximally trained strategies (Evolved Looker Up
1\_1\_1),
which always defects if the opponent defects in the first round, is effectively
employing a one-shot handshake of C. Similarly, Grudger (also known as Grim),
which emerged from training memory one strategies for the Moran process,
also effectively employs a handshake of always cooperating, as it defects
for the remainder of the match if the opponent ever defects.

The insights that payoff maximizers are better invaders and that handshakers
are better resistors suggests that a strategy
aware of the population distribution could choose to become a handshaker at
a critical threshold and use a strategy better for invasion when in the
minority. Information about the population distribution was not available
to our strategies. Previous work has showed that strategies able to retain
memory across matches can infer the population distribution and act in such
a manner, resulting in a strategy effective at invasion and resistance
\cite{Lee2015}.

We did not attempt other objective functions that may serve to select for both
invasion and resistance better than training at a starting population of
$(N/2, N/2)$. Nevertheless our results suggest that there is not much room for
improvement. Any handshake more sophisticated than always cooperate necessarily involves
a defection. (A strategy with a handshake consisting of a long sequence of cooperations is
effectively a grudger.) For TF3 or EvolvedLookerUp1\_1\_1 to become better resistors
they need a longer or more strict handshake. But if this handshake involves
a defection then likely the invasion ability is diminished for $N > 2$: the top
invaders for larger $N$ are nice strategies that do not defect before their
opponents. This is because good invaders need to maximize match payoff to benefit
from fitness proportionate selection,
and so in the absence of a handshake mechanism, knowledge of the population
distribution, or some identifying label on the opponent,
a strategy must be generally cooperative. Aggressive strategies
are only effective invaders for the smallest $N$, dropping dramatically in rank
as the population size increases.

We did, however, attempt to evolve CS using finite state machines and lookup table
based players,
which resulted in some very similar strategies. In particular we evolved a
lookup strategy that had a handshake of DC and played TFT with other players
after a correct handshake while defecting otherwise, which is quite close in
function to CS (full grudging is not possible with a lookup table of limited
depth).

@@

A detailed empirical analysis of 164 strategies of the IPD within a pairwise
Moran process has been carried out. All \(\binom{164}{2}=13,366\) possible
ordered pairs of strategies have been placed in a Moran process with different
starting values allowing the each strategy to attempt to invade the other.
This is the largest such experiment carried out and has led to many insights.

@@

Memory one strategies do not perform as well as longer memory strategies in general
in this study. Several longer memory strategies were high performers for invasion,
particularly the strategies which have been trained using a number of reinforcement
learning algorithms. Interestingly they have been trained to perform well in
tournaments and not Moran processes specifically. In some cases these strategies
utilize all the history of play (the neural network strategies and the lookup
table strategies, the latter using the first round and some number of trailing rounds).

There are no memory one strategies in the top 5 performing strategies
for \(N>3\). Training memory-one strategies specifically for the Moran process
typically led to Grudger / Grim, a memory-one strategy with
four-vector (1, 0, 0, 0). It appears to be the best resistor of the memory-one strategies.
The highest performing memory-one strategy for invasion is PSO Gambler Mem 1,
training to maximize total payout, which has four-vector $(1, 0.52173487, 0, 0.12050939)$.
For comparison, training for maximum score difference between the player and the
opponent resulted in a strategy nearly the same as Grudger, with four-vector
$(0.9459, 0, 0, 0)$ (not included in the study).
